---
title: "GRN inference from scRNAseq data"
author: "Malvina Marku"
date: "10/6/2021"
bibliography: [grn_references.bib]
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is an R Markdown performing GRN inference using different inference methods, as described in [@Kang2021]. Here different methods of GRN inference are used, including GINIE3 [@Huynh2010]. For each method used, a brief description will be given. 

## Methods
Following the comparison analysis of 


## Data processed and results

The data used in this example are taken from the sources described in [@Kang2021]. They consist in scRNAseq data on retina cells, T-cells and hematopoietic cells [@Hay2018] [@Setty2019]. 

Load the necessary libraries for performing the analysis
```{r}
library(GENIE3)
library(beepr) # For notifying when then simulation ends

```

As a first step, we upload the data (**Note:** in the table format, all the rows should correspond to the genes and the columns to the samples.)


<!-- ```{r}  -->
<!-- exprMatr <- matrix(sample(1:10, 100, replace = TRUE), nrow = 20) -->
<!-- rownames(exprMatr) <- paste("Gene", 1:20, sep = "") -->
<!-- colnames(exprMatr) <- paste("Sample", 1:5, sep = "") -->
<!-- head(exprMatr) -->
<!-- ``` -->

```{r}
expression_data <- t(read.csv("HEMATO/Hay_Mono.csv"))
colnames(expression_data) <- paste("Sample", 1:ncol(expression_data), sep = "")
expression_data <- expression_data[-c(1),]
```

We then calculate the weight matrix between the genes using GENIE3: 

```{r}
set.seed(123) # For reproducibility of results
weightMat <- GENIE3(expression_data, nCores = 8, verbose = TRUE)
beep(3) # For notifying when the simulation ends
```

For large datasets, several options can be used to calculate the weight matrix, particularly in restricting the number of regulators per gene, but also for changing the tree-based method settings. For example: 

1. Declare the regulators and apply the weight matrix calculation for this list of regulators only: `regulators <- c("Gene1", "Gene2", "Gene3"), weightMat <- GENIE3(exprMatr, regulators=regulators)`.
2. GENIE3 is based on regression trees. These trees can be learned using either the Random Forest method [@Breiman2001] (`tree.method = "RF"`, default) or the Extra-Trees method [@Geurts2006] (`tree.method = "ET"`). Each tree-based method has two parameters: `K` and `ntrees`. `K` is the number of candidate regulators that are randomly selected at each tree node for the best split determination. The parameter `ntrees` specifies the number of trees that are grown per ensemble. It can be set to any strictly positive integer (the default value is 1000).
3. To decrease the computing times, GENIE3 can be run on multiple cores. The parameter `ncores` specifies the number of cores you want to use. For example:

`set.seed(123) # For reproducibility of results, weightMat <- GENIE3(exprMatr, nCores = 4, verbose = TRUE)`
Note that `seet.seed` allows to get the same results across different runs, but only within `nCores == ` or `nCores > 1`. e.g. A run with `set.seed(123)` and `nCores = 1` and another with the same seed but `nCores > 1` may provide different results.

The list of all the regulatory links (from most likely to least likely) can be obtained with: 

```{r}
linkList <- getLinkList(weightMat)

```

This matrix contains the ranking of all the genes, in which each row correspond to a regulatory link. From here, we can select the top `n` links (e.g. `linkList <- getLinkList(weightMat, reportMax = 10)`), or apply a threshold rank to select the most likely ones (e.g. `linkList <- getLinkList(weightMat, threshold = 0.5)`). 
**Note:** because of the intrinsic randomness of RT or ET methods, the rankings of the links can be different from one run to another. 

Once the network links are obtained, we can apply *filtering*, in order to filter only the `k` most important links. 

```{r}
k = 100000

setwd('../scNET/Results/')

reg_net = read.table('Genie3_Network.tsv', quote = F)

```


## Algorithm reproducibility
There are several metrics we can use to evaluate the reproducibility of the algorithm on 2 different datasets (on the same cells). Among these we mention: *Intersection Index (II)*, *Weighted Jaccard Similarity (WJS)* and *RcisTarget Score (RTS)*. 

```{r}
Rcis_GENIE3 = Custom.Rcis(input.dir = 'scNET/Results',
                     pattern = 'GENIE3',
                     chosenDb="hg19-tss-centered-5kb-7species.mc9nr.feather",
                     output.dir = 'scNET/Results',
                     MinGenesetSize = 0, 
                     directed = T)
Results= Reproducibility.stats('Genie3*.tsv'),
                               Results.dir = 'scNET/Results')

Results[,'RcisTarget_index'] = Rcis.percent(Rcis_GENIE3)

Results
```

