---
title: "GRN Inference from CLL time-course bulk RNA-seq"
author: "Malvina Marku, Hugo Chenel, Flavien Raynal"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
editor_options: 
  chunk_output_type: console
---

This is the main code used for inferring the GRN for time-series of bulk RNAseq data of in-vitro monocyte-CLL experiment. Here we use the time-series of the CLL cells, in dataset of 11-13 time points. We use dynGENIE3 for inferring the regulatory network.

Work in progress. 

**1. Load the libraries and `dynGENIE3` functions:**

```{r include=FALSE}
library(reshape2)
library(doRNG)
library(doParallel)
library(tidyverse)
library(DESeq2)
library(clusterSim)
library(dplyr)
library(ggplot2)
library(CINNA)
library(igraph)
library(qgraph)

source("dynGENIE3.R")
```

**2. Load the data**
In our case, we have the time-course transcriptomics from 2 patients and 2 replicates per patient. the inference method will be applied in different ways:
\item patient 1 and 2 taken separately. The average between replicates will be calculated per patient.
\item ...

```{r include=FALSE}
Patient_1_rep_1_table <- read.expr.matrix("/mnt/SERVER-CRCT-STORAGE/CRCT21/Private/CLL_dec2020/RNAseqBulk_Eq21_Eq9/Gene_counts_tables/Gene_counts_Patient1_rep1_TPM.txt",form = "rows.are.genes")

Patient_1_rep_2_table <- read.expr.matrix("/mnt/SERVER-CRCT-STORAGE/CRCT21/Private/CLL_dec2020/RNAseqBulk_Eq21_Eq9/Gene_counts_tables/Gene_counts_Patient1_rep2_TPM.txt", form = "rows.are.genes")

Patient_2_rep_1_table <- read.expr.matrix("/mnt/SERVER-CRCT-STORAGE/CRCT21/Private/CLL_dec2020/RNAseqBulk_Eq21_Eq9/Gene_counts_tables/Gene_counts_Patient2_rep1_TPM.txt", form = "rows.are.genes")

Patient_2_rep_2_table <- read.expr.matrix("/mnt/SERVER-CRCT-STORAGE/CRCT21/Private/CLL_dec2020/RNAseqBulk_Eq21_Eq9/Gene_counts_tables/Gene_counts_Patient2_rep2_TPM.txt", form = "rows.are.genes")
```


```{r}
TS1 = read.expr.matrix("Data/time_series_1.txt", form="rows.are.samples")
TS2 = read.expr.matrix("Data/time_series_2.txt", form="rows.are.samples")
TS3 = read.expr.matrix("Data/time_series_3.txt", form="rows.are.samples")
```


**Re-format the data:**

_Firstly_, make sure there is no `NA, NaN, or Inf` value in the dataset: 

```{r include=FALSE}
#Identify the NA, NaN and Inf values
impute.mean <- function(x) replace(x, is.na(x) | is.nan(x) | is.infinite(x) | is.factor(x), mean(x, na.rm = TRUE))

Patient_1_rep_1_table <- apply(Patient_1_rep_1_table, 2, impute.mean)
sum( apply( Patient_1_rep_1_table, 2, function(.) sum(is.infinite(.))) )

Patient_1_rep_2_table <- apply(Patient_1_rep_2_table, 2, impute.mean)
sum( apply( Patient_1_rep_2_table, 2, function(.) sum(is.infinite(.))) )


```


```{r}
time.points = list(TS1[1,], TS2[1,], TS3[1,])
TS.data = list(TS1[2:nrow(TS1),], TS2[2:nrow(TS2),], TS3[2:nrow(TS3),])
```



_Secondly_, re-format the time-series:

```{r include=FALSE}
#Add a row in each of the datasets indicating the time point

Patient_1_time_points <- c(1:ncol(Patient_1_rep_1_table))
Patient_2_time_points <- c(1:ncol(Patient_2_rep_1_table))

#Remove the rows with 0 variance
Patient_1_rep_1_table <- rbind(Patient_1_time_points, Patient_1_rep_1_table)
Patient_1_rep_1_table <- Patient_1_rep_1_table[rowSums(Patient_1_rep_1_table[]) > 0,]

Patient_1_rep_2_table <- rbind(Patient_1_time_points, Patient_1_rep_2_table)
Patient_1_rep_2_table <- Patient_1_rep_2_table[rowSums(Patient_1_rep_2_table[]) > 0,]

Patient_2_rep_1_table <- rbind(Patient_2_time_points, Patient_2_rep_1_table)
Patient_2_rep_1_table <- Patient_2_rep_1_table[rowSums(Patient_2_rep_1_table[]) > 0,]

Patient_2_rep_2_table <- rbind(Patient_2_time_points, Patient_2_rep_2_table)
Patient_2_rep_2_table <- Patient_2_rep_2_table[rowSums(Patient_2_rep_2_table[]) > 0,]

# write.table(Patient_1_rep_1_table, file = "Patient_1_rep_1_table.txt")
# write.table(Patient_1_rep_2_table, file = "Patient_1_rep_2_table.txt")
# write.table(Patient_2_rep_1_table, file = "Patient_2_rep_1_table.txt")
# write.table(Patient_2_rep_2_table, file = "Patient_2_rep_2_table.txt")

# Patient_1_rep_1_table <- read.expr.matrix("Patient_1_rep_1_table.txt",form = "rows.are.genes")
# Patient_1_rep_2_table <- read.expr.matrix("Patient_1_rep_2_table.txt", form = "rows.are.genes")
# Patient_2_rep_1_table <- read.expr.matrix("Patient_2_rep_1_table.txt", form = "rows.are.genes")
# Patient_2_rep_2_table <- read.expr.matrix("Patient_2_rep_2_table.txt", form = "rows.are.genes")

### Normalize data in [0, 1] -- as in the example model
### Normalization method: n4 (https://rdrr.io/cran/clusterSim/man/data.Normalization.html)
#Patient_1_rep_1_unit = as.data.frame(data.Normalization(Patient_1_rep_1_table, type = "n4", normalization = "column" ))
#Patient_1_rep_2_unit = as.data.frame(data.Normalization(Patient_1_rep_2_table, type = "n4", normalization = "column" ))
#Patient_2_rep_1_unit = as.data.frame(data.Normalization(Patient_2_rep_1_table, type = "n4", normalization = "column" ))
#Patient_2_rep_2_unit = as.data.frame(data.Normalization(Patient_2_rep_2_table, type = "n4", normalization = "column" ))

#Find genes that appear in both df
common_genes_1 <- intersect(rownames(Patient_1_rep_1_table), rownames(Patient_1_rep_2_table))
common_genes_2 <- intersect(rownames(Patient_2_rep_1_table), rownames(Patient_2_rep_2_table))
common_genes <- intersect(common_genes_1, common_genes_2)

#Keep only the genes that appear in common_genes
Patient_1_rep_1_filter <- as.matrix(subset(Patient_1_rep_1_table, rownames(Patient_1_rep_1_table) %in% common_genes))
Patient_1_rep_2_filter <- as.matrix(subset(Patient_1_rep_2_table, rownames(Patient_1_rep_2_table) %in% common_genes))
Patient_2_rep_1_filter <- as.matrix(subset(Patient_2_rep_1_table, rownames(Patient_2_rep_1_table) %in% common_genes))
Patient_2_rep_2_filter <- as.matrix(subset(Patient_2_rep_2_table, rownames(Patient_2_rep_2_table) %in% common_genes))

#p1_rep1 = do.call(rbind.data.frame, time_series_data[1])
#p1_rep2 = do.call(rbind.data.frame, time_series_data[2])
#p2_rep1 = do.call(rbind.data.frame, time_series_data[3])
#p2_rep2 = do.call(rbind.data.frame, time_series_data[4])

time_series_data <- list(Patient_1_rep_1_filter[1:nrow(Patient_1_rep_1_filter),], Patient_1_rep_2_filter[1:nrow(Patient_1_rep_2_filter),], Patient_2_rep_1_filter[1:nrow(Patient_2_rep_1_filter),], Patient_2_rep_2_filter[1:nrow(Patient_2_rep_2_filter),])

df_time_series_data = as.data.frame(time_series_data)

sample_info = as.data.frame(str_split_fixed(colnames(df_time_series_data),pattern = "_", n = 3))
names(sample_info)[names(sample_info) == "V1"] <- "Day"
names(sample_info)[names(sample_info) == "V2"] <- "Patient"
names(sample_info)[names(sample_info) == "V3"] <- "Replicate"

#Create DESeq2 object
dds <- DESeqDataSetFromMatrix(countData = round(df_time_series_data), colData = sample_info, design = ~ Day + Patient + Replicate)
nrow(dds) #16432
#Remove rows having low counts
firstcondition <- rowSums(counts(dds) > 1) >= 2 #at least 2 samples with a count of 2 or higher
dds <- dds[firstcondition,]
nrow(dds) #12247

#Generate normalized counts
dds <- estimateSizeFactors(dds)
sizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE)

#Remove all genes with zero variance (i.e constant expression)
variances = rowVars(normalized_counts)
range(variances)
normalized_counts = normalized_counts[!rowVars(normalized_counts) == 0 ,]

#Keep only rows corresponding to TF
normalized_counts = as.data.frame(normalized_counts)
TFs_df = read.csv("Data/TF_names_v_1.01.txt", header = FALSE)
TFs_vector = TFs_df[['V1']]
TFs_vector = TFs_vector[TFs_vector %in% rownames(normalized_counts)] 
normalized_counts_TF <- normalized_counts[rownames(normalized_counts) %in% TFs_vector, ]


#List of matrices corresponding to the different patients/replicates 
df_conditions = sapply(c("p1_rep1", "p1_rep2", "p2_rep1", "p2_rep2"),
                function(x) normalized_counts_TF[endsWith(names(normalized_counts_TF),x)],
                simplify = FALSE)

df_conditions$p1_rep1 = as.matrix(df_conditions$p1_rep1)
df_conditions$p1_rep2 = as.matrix(df_conditions$p1_rep2)
df_conditions$p2_rep1 = as.matrix(df_conditions$p2_rep1)
df_conditions$p2_rep2 = as.matrix(df_conditions$p2_rep2)
normalized_counts_list <- list(df_conditions$p1_rep1[1:nrow(df_conditions$p1_rep1),], df_conditions$p1_rep2[1:nrow(df_conditions$p1_rep2),], df_conditions$p2_rep1[1:nrow(df_conditions$p2_rep1),], df_conditions$p2_rep2[1:nrow(df_conditions$p2_rep2),])


# Create the time points
time.points_1 <- c(1,2,3,4,5,6,7,8,9,11,13)
time.points_2 <- c(1,2,3,4,5,6,7,8,9,10,11,12,13)

time.points <- list(time.points_1, time.points_1, time.points_2, time.points_2)
#time.points <- list(Patient_1_rep_1_filter[1,], Patient_1_rep_2_filter[1,], Patient_2_rep_1_filter[1,], Patient_2_rep_2_filter[1,])

decay_rates = 0.2

write.table(normalized_counts, file="Results/normalized_counts.txt", sep="\t", quote=F, col.names=NA)
write.csv(normalized_counts_list,"Results/normalized_counts_TF_list.csv", row.names = TRUE)
```


**Run `dynGENIE3` with default parameters or by tuning the parameters:**

```{r eval=FALSE, include=FALSE}
run_analysis <- dynGENIE3(normalized_counts_list, time.points, alpha = decay_rates, tree.method = "RF", SS.data=NULL, K = "sqrt", ntrees = 1000, regulators = TFs_vector, ncores = 12, verbose = FALSE, seed = NULL)
```

**Filter and visualize the results**
First, we plot the distribution of link.list, to help select a threshold.

```{r eval=FALSE, include=FALSE}
link.list <- get.link.list(run_analysis$weight.matrix)
links_filtered_TF <- filter(link.list, link.list$weight >= 0.01) #3749 links
write.csv(links_filtered_TF, "Results/links_filtered_TF.csv", row.names = T, col.names = T, sep = " ")

ggplot(link.list, aes(x=weight)) + 
  geom_histogram(color="black", fill="white") +
  labs(title="Links weight distribution", subtitle="Edges weight histogram plot") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), plot.subtitle = element_text(hjust = 0.5)) +
  xlab("Weight")+ylab("Count")

ggplot(links_filtered_TF, aes(x=weight)) + 
  geom_histogram(color="black", fill="white") +
  xlim(min(links_filtered_TF$weight), max(links_filtered_TF$weight)) +
  labs(title="Links filtered weight distribution", subtitle="Edges weight histogram plot") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), plot.subtitle = element_text(hjust = 0.5)) +
  xlab("Weight")+ylab("Count") + geom_vline(xintercept= min(links_filtered_TF$weight), size = 2, colour = "red")
```


**Filter on link number**

```{r}
reg_gene = table(links_filtered_TF$regulatory.gene) 
tar_gene = table(links_filtered_TF$target.gene)
frequency_reg_gene = as.data.frame(reg_gene) #correspond to the outdegree of a node
frequency_tar_gene = as.data.frame(tar_gene) #correspond to the indegree of a node

df_gene_degree = frequency_reg_gene %>% full_join(frequency_tar_gene,by="Var1")

df_gene_degree = df_gene_degree %>%
  rowwise() %>% 
  mutate(f_new=sum(Freq.x, Freq.y, na.rm = T)) #sum of frequencies for regulatory genes, target genes and both of them (=gene degree)

df_gene_degree = df_gene_degree[,-(2:3),drop=FALSE]
df_gene_degree = df_gene_degree[order(df_gene_degree$Var1), ]

names(df_gene_degree)[names(df_gene_degree) == "Var1"] <- "Gene"
names(df_gene_degree)[names(df_gene_degree) == "f_new"] <- "Gene.degree"

df_gene_degree = df_gene_degree %>%
                   arrange(desc(Gene.degree))

#Gene degree distribution
ggplot(df_gene_degree, aes(x=Gene.degree)) + 
    geom_histogram(color="black", fill="white", bins=100) +       
    ggtitle("Gene degree distribution") + xlab('Gene degree') + ylab('Frequency') +
    theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))
ggplot(df_gene_degree, aes(x=Gene.degree)) + 
    geom_histogram(color="black", fill="white", bins=100) +         
    xlim(-10, 100) + ggtitle("Gene degree distribution for genes with the lowest gene degree") + xlab('Gene degree') + ylab('Frequency') + theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))
  
#filter on genes according to their gene degree
df_gene_degree_filtered = filter(df_gene_degree, Gene.degree >= 3)
df_gene_degree_removed = filter(df_gene_degree, Gene.degree < 3)

#remove all the links where gene degree < 3
links_filtered_TF_treshold = links_filtered_TF[ ! (links_filtered_TF$regulatory.gene %in% df_gene_degree_removed$Gene) & ! (links_filtered_TF$target.gene %in% df_gene_degree_removed$Gene), ]

write.csv(df_gene_degree,"/home/crct2112/Documents/GRN/Results/TF/Gene_degree.csv", row.names = TRUE)
write.csv(links_filtered_TF_treshold,"/home/crct2112/Documents/GRN/Results/TF/Links_filtered_treshold.csv", row.names = TRUE, quote = FALSE) #ne plus utiliser celui là
```


**CINNA**

```{r}
g1 <- graph.data.frame(links_filtered_TF, directed = TRUE) #create graph from datatable
```

```{r}
vcount(g1) #1004
ecount(g1) #3749
plot(g1)
g1
```

```{r}
#Make sure there is no multiple edges or loops
g1s <- simplify(g1, remove.multiple = T, remove.loops = T, 
                 edge.attr.comb=c(weight="sum", type="ignore"))
vcount(g1s)
ecount(g1s)
```



##Network components analysis

###Giant component extraction
```{r}
g2 = giant_component_extract(g1, directed = TRUE) #extract the largest connected component of the Graph
giant_comp = g2[[1]]

vcount(g1) #1004
ecount(g1) #3749
plot(g1)

vcount(giant_comp) #1002
ecount(giant_comp) #3748
plot(giant_comp)

g = as_data_frame(giant_comp)
names(g)[names(g) == "from"] <- "Regulatory Gene"
names(g)[names(g) == "to"] <- "Target Gene"
write.csv(g,"Results/Giant_comp_0.01.csv", row.names = TRUE, quote = FALSE)
```



**Graph creation from datatable and exploration**

```{r}
g0 = graph.data.frame(links_filtered_TF, directed = TRUE) 
g1 = graph.data.frame(g, directed = TRUE)

#Size distribution of connected components
clu0 = components(g0)
clu0$no #Number of distinct cluster
clu0$csize #Size of clusters

ggplot() + aes(clu0$csize) + 
  geom_histogram(binwidth=100, colour="black", fill="white") + 
  labs(title="Cluster size distribution", subtitle="Histogram of size of connected components", caption="Node list : 1004 \n Edge list : 3749") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"), plot.subtitle = element_text(hjust = 0.5), plot.caption = element_text(face = "italic", color = "gray40")) +
  xlab("Cluster size") + ylab("Frequency")

clu1 = components(g1)
clu1$no
clu1$csize #one connected component of 1002 nodes
```


**Centrality measure analysis**

###Suggestion of proper centralities based on network topology
```{r}
proper_centralities(zachary) #Undirected and unweighted network
proper_centralities(rhesus) #Directed and weighted network
proper_centralities(g1)
```

###Centrality computations
```{r}
calculate_centralities(giant_comp, include = "Degree Centrality")
pr_cent = proper_centralities(g1)
```

###Best centrality measures
```{r}
calc_cent = calculate_centralities(g1, include = pr_cent[1:3])
calc_cent
```

###Recognition of most informative measures
```{r}
pca_centralities(calc_cent) #Page.Rank, Burt.s.Constraint and Alpha.Centrality
pca_centralities(calc_cent, scale.unit = FALSE)

tsne_centralities(calc_cent, dims = 2, perplexity = 0.5, scale=TRUE) #Page.Rank, Burt.s.Constraint and Alpha.Centrality
```


##Visualization on centrality analysis

###Graph visualization regarding to the centrality type
```{r}
#Correlation between computed centrality measures
visualize_correlations(calc_cent, "pearson")

#Regression across centrality measures
Alpha_centrality = calc_cent[[1]] 
Burt_constraint = calc_cent[[2]]
visualize_association(Alpha_centrality, Burt_constraint)

PageRank = calc_cent[[3]]
visualize_association(PageRank, Alpha_centrality )

#Pairwize correlation between centrality types
visualize_pair_correlation(Alpha_centrality, Burt_constraint) #negative correlation between Alpha_centrality and Burt_constraint
visualize_pair_correlation(PageRank, Alpha_centrality) #positive correlation between PageRank and Alpha_centrality

```


**Graph exploration**

## 1 - Degree centrality (Local Centrality Measures)
```{r}
g1_deg = degree(g1) #Degree centrality (gene degree)
g1_in_deg = degree(g1, mode="in")
g1_out_deg = degree(g1, mode="out")
g1_deg_distr = degree_distribution(g1)
V(g1)$Degree = g1_deg
which.max(g1_deg) #ZNF718
```

## 2 - Closeness centrality (Global Centrality Measures)
```{r}
g1_closeness = closeness(g1, mode="all")
V(g1)$Closeness = g1_closeness
which.max(g1_closeness) #ETS1
```


## 3 - Eigenvector centrality
```{r}
g1_evcent = eigen_centrality(g1)$vector
V(g1)$Eigen = g1_evcent
which.max(g1_evcent) #ZNF718
```


## 4 - Betweenness centrality
```{r}
g1_bw = betweenness(g1)
g1_ebw = edge_betweenness(g1) #might give false values for graphs with multiple edges. 
V(g1)$Betweeness = g1_bw
which.max(g1_bw) #BHLHE40
```

## 5 - PageRank
```{r}
g1_pr = page_rank(g1)$vector
V(g1)$PageRank = g1_pr
which.max(g1_pr) #ZNF667
```

## 6 - Burt's constraint
```{r}
g1_ct = constraint(g1)
V(g1)$Constraint = g1_ct
#V(g1)$Constraint
which.max(g1_ct) #ZNF519
```

## 7 - Alpha centrality
```{r}
g1_ac = alpha_centrality(g1)
V(g1)$Alpha = g1_ac
which.max(g1_ac) #VENTX
```


##Convert iGraph object into dataframe 
```{r}
g1_df = as_long_data_frame(g1)
write.csv(g1_df,"Results/Graph_data_info.csv", row.names = TRUE)
```



#Graph optimization

Remove zinc finger (with high gene degree) and look at the connectivity of the network
 
```{r}

which.max(g1_deg) #ZNF718
log_degree_g1 = as.data.frame(log(degree(g1)))

ggplot((log_degree_g1), aes(x=log(degree(g1)))) + 
  geom_histogram(color="black", fill="white") +
  labs(title="Gene degree distribution") +
  theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
  xlab("Gene degree (log)") + ylab("Frequency")

sort(degree(g1), decreasing = TRUE)[1:30]
g1_selected = delete_vertices(g1,c("ZNF718","ZNF675", "ZNF706")) #remove zinc fingers with the highest gene degree
g1_selected #999 nodes and 3606 edges
sort(degree(g1_selected), decreasing = TRUE)[1:20]

clug1 <- components(g1_selected)
groups(clug1)
clug1$no #single cluster for the whole network
clug1$csize #cluster of 999 already connected

# Identify vertices with gene degree <= threshold
degree(g1_selected, mode = "out")
table(degree(g1_selected, mode = "out")) #distribution of nodes according to gene degree
#remove leafs : 165 (nodes with zero outdegree)
remove <- which(degree(g1_selected, mode = "out") < 1)
# Delete vertices and associated edges
giant_comp_selected_leaffree <- delete.vertices(g1_selected, remove)

table(degree(giant_comp_selected_leaffree, mode = "out"))
vcount(giant_comp_selected_leaffree) #834
ecount(giant_comp_selected_leaffree) #3054

g = igraph::as_data_frame(giant_comp_selected_leaffree)
names(g)[names(g) == "from"] <- "Regulatory Gene"
names(g)[names(g) == "to"] <- "Target Gene"
write.csv(g,"Results/Giant_comp_0.01_selected.csv", row.names = TRUE, quote = FALSE)

```



