---
title: "Gene Ontology"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


############################# GENE ONTOLOGY #############################


### Libraries loading ###

```{r}
library(biomaRt)
library(limma)
library(stringr)
library(GO.db)
library(org.Hs.eg.db)
library(clusterProfiler)
library(MOMA)
library(dplyr)
library(tidyverse)
library(DOSE)
library(enrichplot)
library(clusterProfiler)
library(ggupset)
```


### Gene list ###

```{r}
TF_list = read.csv("Results/TF_list.csv")
mygenes = as.character(TF_list$x)
length(mygenes) #140

TF_total = read.csv("Data/TF_names_v_1.01.txt", header = FALSE)
TF_total = as.character(TF_total$V1)
```


### Modify gene list to add first neighbours or targets (facultative) ###

```{r}
#Explain how to obtain TF and their targets (dynGENIE3)
TF_and_targets <- read.csv("/home/crct2112/Documents/GRN/Results/TF/Links_filtered_TF.csv")
TF_and_targets <- dplyr::select(TF_and_targets, -c(X))

TF_list = read.csv("Results/TF_list.csv")
TF_list = dplyr::select(TF_list, -c(X))
names(TF_list)[names(TF_list) == "x"] <- "regulatory.gene"

TF_gene_list = merge(TF_list, TF_and_targets, by = "regulatory.gene", all.x=TRUE, all.y=FALSE)
TF_gene_list = dplyr::select(TF_gene_list, -c(weight))

TF_gene_list = TF_gene_list %>% drop_na()

TF_gene_list2 = as.data.frame(rbind(as.matrix(TF_gene_list$regulatory.gene), as.matrix(TF_gene_list$target.gene)))
TF_gene_list2 = distinct(TF_gene_list2,V1, .keep_all= TRUE) #remove duplicates
mygenes = as.character(TF_gene_list2$V1)
length(mygenes) #4855 genes
```
This list contains all the TFs of the network and their targets.

### Format conversion ###

```{r}
#Convert from hugo gene names to entrez ids

#mapHugo(c("ADNP","ADNP2","AEBP1"))
TF_total_ID = mapHugo(TF_total)
```

### Remove version from ensembl id ###

```{r}
#mygenes=str_split_fixed(mygenes,"\\.",2)[,1]
```

### ID conversion Ensembl to Entrez ###

```{r}
# library("TxDb.Hsapiens.UCSC.hg38.knownGene")
# geneHg38.gr <- genes(TxDb.Hsapiens.UCSC.hg38.knownGene,columns="gene_id")
# gene_id <- geneHg38.gr$gene_id
# orgDBgenes <- select(org.Hs.eg.db,keys = gene_id,columns = "SYMBOL",keytype = "ENTREZID")
# geneHg38.gr$SYMBOL <- orgDBgenes[match(gene_id,orgDBgenes[,1]),2]
# myentrez=geneHg38.gr$gene_id[which(geneHg38.gr$SYMBOL %in% mygenes)]
```


### BioMart ###

```{r}
ensembl = useEnsembl(biomart="genes")
datasets <- listDatasets(ensembl)
datasets
ensembl = useEnsembl(biomart="genes", dataset="hsapiens_gene_ensembl")

listAttributes(ensembl)
genes.with.id=getBM(attributes=c("hgnc_symbol", "entrezgene_id"), values=mygenes, mart=ensembl)

id = genes.with.id[which(genes.with.id$hgnc_symbol %in% mygenes),]
nrow(id)
myentrez = unique(id$entrezgene)

length(mygenes) #140 genes
length(myentrez) #133 genes
```


### Gene Ontology ###
```{r}
goana = goana(de=myentrez, geneid=myentrez, FDR=0.95, species="Hs")#, universe = TF_total_ID)
View(goana)
```


#GO Plot
```{r}
compGO <- compareCluster(geneCluster     = list(genes=myentrez),
                         fun           = "enrichGO",
                         OrgDb         = 'org.Hs.eg.db',
                         ont           = 'BP',
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH",
                         universe = TF_total_ID)
dotplot(compGO, showCategory = 10, title = "Gene Ontology Enrichment Analysis")
```


### KEGG Ontology ###
```{r}
kegga = kegga(de=myentrez, geneid=myentrez, FDR=0.95, species="Hs", universe = TF_total_ID)
View(kegga)
```


#KEGG Plot
```{r}
compKEGG <- compareCluster(geneCluster   = list(genes=myentrez),
                           fun           = "enrichKEGG",
                           pvalueCutoff  = 0.25,
                           pAdjustMethod = "BH",
                           universe = TF_total_ID) 
dotplot(compKEGG, showCategory = 15, title = "KEGG Pathway Enrichment Analysis")
```


### Keep only "BP" ontology and pvalue < 0.05 then sort by pvalue ###
```{r}
goana = goana[which(goana$Ont == "BP" & goana$P.DE < 0.05),]
goana = goana[order(goana$P.DE),]
kegga = kegga[order(kegga$P.DE),]
```


## Which genes are concerned by a GO ID
```{r}
genes.with.id2 = getBM(attributes=c("ensembl_gene_id","hgnc_symbol","go_id", "description"), values=mygenes, mart=ensembl)
id2 = genes.with.id2[which(genes.with.id2$hgnc_symbol %in% mygenes),]

id2[which(id2$go_id=="GO:0016779"),c(2,4)]
```


### Export to csv file ###
```{r}
write.csv(goana, "Results/Gene_Ontology.csv", quote=F)
write.csv(kegga, "Results/KEGG_Pathway.csv", quote=F)
```


### cnet plot ### 
```{r}
x = enrichGO(myentrez, OrgDb = "org.Hs.eg.db", ont = "BP", pvalueCutoff = 0.05, pAdjustMethod = "BH", universe = TF_total_ID)
edox <- setReadable(x, 'org.Hs.eg.db', 'ENTREZID')

cnetplot(edox, categorySize="pvalue", foldChange=myentrez)
cnetplot(edox, foldChange=myentrez, circular = TRUE, colorEdge = TRUE) 
#cowplot::plot_grid(p1, p2, ncol=2, labels=LETTERS[1:2])
upsetplot(edox)
#heatplot(edox)
```


### Tree plot ###
```{r}
edoxT <- pairwise_termsim(edox)
treeplot(edoxT)
```


### Enrichment Map ###
```{r}
edoxE <- pairwise_termsim(edox)
emapplot(edoxE, cex_category=1.7)
```

